\thispagestyle{plain}
\section*{Acknowledgments}
In the six months that brought to this thesis I received a great deal of support from many people, and even before that experiences I shared with many people led me to study these subjects.
First, I'd like to thank my supervisors, Prof. Bruni and Prof. Gori, for the guidance and the help. They pointed me in the right direction at the beginning, provided interesting cues (without which substantial parts of this work wouldn't exist) and reviewed the final result, suggesting useful improvements.

At personal level, I would like to thank a lot of people, but right now I don't think I can remember them all. I apologize to people I'm forgetting to name, but you really are a lot.
I thank my family, my parents and my brother, because they were always supportive of my studies and my choices.
I thank my friends from the university years, Alessandro, Beppe, Bruno, Elia, Enrico, Guido, Niccolò, Pasquale e Sofia: with you these years, including partly this last one, flown by.
I thank my friends from the d\&d party, Dario, Gibbo, Gozzi, Manuel, Marco, Mors, Palla, Tara: I don't know how you felt about it, but in time of pandemic those weekly meetings, even though remotely, really meant a lot for me.
I want to thank all the people I got to know through PCW. I can't list them all because they really are a lot, but I would never have imagined that small edit in 2013 would have become an important part of my life.
I thank my other group of friends from the high-school years, Alge, Boss, Dario, Fra, Gozzi, Gulia, Miki, Nico, Otti, Turlea: your friendship really is dear to me.
Lastly, even though I already mentioned many of them for other reason, I'd like to thank again all the people with whom I participated in high school maths competitions, specially Gozzi and Miki, and the teacher responsible of the project, prof. Greppi: I don't think is an overstatement to say that experience was the first step towards what I study (and enjoy studying!) now, in addition to making me know many of the people I now call friends.